---
title: "Analysing East / West Correlation within Covid"
description: A statistical review of East/West correlation claims for Covid
date: January 24, 2021
output:
  distill::distill_article:
    toc: true
    theme: theme.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggthemr)

# ggthemr::ggthemr()

knitr::opts_chunk$set(echo = F,
                      warning = T,
                      message = T)

```

Recent analysis came out claiming that there was evidence of a East to West pattern of Covid impacts for Europe. As described in the tweet:

https://twitter.com/RealJoelSmalley/status/1352606263838973953/photo/1

The purpose of this article is to provide a statistical review of the claims made, and to help provide clearer explanation around the work. It hopes to provide a more detailed explainer of the original methods, and guide people who may not be familiar with statistical processes through the steps which we can help us be confident in the interpretations.

If this looks detailed, **that is the very point**. I want to explain the decisions along the way which are being made and will aim to be as objective as possible.

# What does the original study claims

> "COVID mortality follows a distinct pattern across Europe, from West to East, apparently determined almost entirely by physical geography and season. There is, however, no apparent correlation between severity of policy intervention and mortality."

> "The different time intervals to peak during epidemic and endemic phases are also consistent in countries that suffered significant COVID exposure in Spring."

> "The similarity in distribution of COVID deaths in countries that share the same longitude is striking. Where I have done the analysis, it fits the profile of usual seasonal mortality."

# Recreating the analysis

To ensure a fair comparison, it is important we first recreate the analysis. As no methodology was published, we are going to try and get as much information out of the tweets as possible. **All this process here is documented for full transparency**. If the assumptions are wrong, they can be tweaked.

## Data Sources

Two data sources were provided:

- The severity index scores the level of interventions provides a measure of the how much restriction is in place to control Covid. More is explained [here](https://www.bsg.ox.ac.uk/research/research-projects/coronavirus-government-response-tracker).
- Weekly Covid deaths per country, as provided by the [EU Open Data Portal(https://data.europa.eu/euodp/en/data/dataset/covid-19-coronavirus-data-weekly-from-17-december-2020)

```{r, message = F}
df <- read_csv("../data/dataCombined.csv") %>%
  mutate(Date = lubridate::ymd(Date))

boundaries <- sf::read_sf("../data/eu_boundaries.gpkg")
```

### Time Periods

As the analysis did not provide a definition from the time periods used for Spring and Winter. The standard definitions for Spring and Winter do not capture the data great (winter technically starts on the 21st December), so instead we will use:

- Spring from start of March until 1st June
- Winter from October until present

```{r}
df <- df %>%
  mutate(Period = ifelse(Date < "2020-06-01", "Spring",
                         ifelse(Date < "2020-10-01", "Summer",
                                "Winter")))
```

This kind of merges Summer and Autumn into one season, but it is not considered too important within the context of the analysis.

## Summary Map

Okay, the article starts with a summary of the deaths in Winter and Spring. This recreation is shown below. The colour scale is a bit off, but you can see the point! You'll notice we have a few more maps.

```{r}
df_summary <- df %>%
  group_by(CountryName, Period) %>%
  summarise(total = sum(deaths_weekly, na.rm = T)) %>%
  ungroup() %>%
  filter(Period != "Summer")

boundaries_summary <- boundaries %>%
  right_join(df_summary, by = c("CountryName"))

ggplot(boundaries_summary) +
  geom_sf(aes(fill = total)) +
  facet_wrap(vars(Period)) +
  coord_sf(xlim = c(-11, 30), ylim = c(34, 60)) +
  scale_fill_gradient(low = "#EDBDAF", high = "#675273") +
  labs(title = "COVID Mortality, Europe", 
       fill = "Deaths In Period",
       caption = "Based off original map produced by Joel Smalley") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        panel.grid = element_blank(),
        plot.background = element_rect(fill = "#FAF7F2"),
        legend.background = element_blank())
```
## Plots

Again, we can recreate the plots. Before we jump into the grouping, I am going to present them as a single list. It is useful to make a quick comparison yourself to see whether you notice there are any potential similarities in the dataset:

```{r, fig.height=12, fig.width=6}
countries_select <- c("United Kingdom", "Ireland", "Portugal", "France", "Spain", "Belgium", "Netherlands", "Italy", "Germany", "Switzerland", "Austria", "Slovenia", "Croatia", "Slovakia", "Poland", "Czech Republic", "Bosnia and Herzegovina")

df_selected <- df %>%
  filter(CountryName %in% countries_select)

maxAxis <- max(df_selected$deaths_weekly) * 1.05

ggplot(df_selected, aes(x = Date)) +
  geom_col(aes(y = deaths_weekly)) +
  facet_wrap(vars(CountryName), ncol = 1, scales = "free_y") +
  labs(y = "Deaths Weekly",
       caption = "Data from EU Open Data Portal")
```

Having had a look without any splitting, we can now plot these separately:

```{r, fig.height=10, fig.width=6}
countries_select <- c("United Kingdom", "Ireland", "Portugal", "France", "Spain", "Belgium", "Netherlands", "Italy", "Germany", "Switzerland", "Austria")
countries_df <- data.frame(CountryName = factor(countries_select, levels = countries_select),
                           Group = as.factor(c(1, 1, 1, 2, 2, 2, 3, 3, 4, 4, 4)))


df_selected <- df %>%
  filter(CountryName %in% countries_select) %>%
  left_join(countries_df) %>%
  mutate(CountryName = factor(CountryName, levels = countries_select))


ggplot(df_selected, aes(x = Date)) +
  geom_col(aes(y = deaths_weekly, fill = Group)) +
  facet_wrap(vars(CountryName), ncol = 1, scales = "free_y") +
  labs(y = "Deaths Weekly",
       title = "Weekly Covid deaths per Country",
       subtitle = "Grouped as defined by Joel Smalley") 
```

# Times Series Clustering

Okay, so hopefully all of the work so far has made sense! This is where things risk getting a little more complicated. A potential problem with the first approach is that it relies a lot on human intuition. We are comparing time series and it at risk of being subjective.

As an alternative way

dtwclust

partition clustering. The aim is to create a similar number of groups as Joel.

https://cran.r-project.org/web/packages/dtwclust/dtwclust.pdf

```{r}
library(dtwclust)
library(imputeTS)

countries_select <- boundaries$CountryName

df_selected <- df %>%
  filter(CountryName %in% countries_select) %>%
  mutate(CountryName = factor(CountryName, levels = countries_select))


df_selected_wide <- 
  df_selected %>%
  select(c(CountryName, Date, deaths_weekly)) %>%
  mutate(deaths_weekly = replace_na(deaths_weekly, 0)) %>%
  arrange(Date) %>%
  pivot_wider(values_from = "deaths_weekly", names_from = "Date") %>%
  column_to_rownames("CountryName")


df_selected_wide_ts <- 
  df_selected_wide %>%
  ts() %>%
  na_interpolation()

# Load series
model <- tsclust(df_selected_wide_ts, type = "partitional",  distance = "dtw_basic", centroid = "pam",  k = 6, trace = T)


plot(model)
```



```{r}
results <- data.frame(CountryName = rownames(df_selected_wide),
                      Cluster = predict(model, df_selected_wide_ts))

boundaries_results <- 
  boundaries %>%
  left_join(results)

ggplot(boundaries_results) +
  geom_sf(aes(fill = as.factor(Cluster))) +
  coord_sf(xlim = c(-11, 30), ylim = c(34, 60)) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        panel.grid = element_blank(),
        plot.background = element_rect(fill = "#FAF7F2"),
        legend.background = element_blank())
```

```{r}
library(spdep)

```





```{r}
knitr::knit_exit()
```








# Classifying Our Countries

In the post, 

- **Significant peaks in March**: **"different time intervals to peak during epidemic and endemic phases are also consistent in countries that suffered significant COVID exposure in Spring."**
- **Countries with no peak in March**:

In order to more rigorously assess the suitability of his claims, we need to first find a way of classifying the rules which can be used to classify the countries in line with his categories

- 






```{r fig.height=10, fig.width=6}


boundaries_summary <- boundaries %>% 
  mutate(studied = ifelse(CountryName %in% countries_select, TRUE, FALSE))

ggplot(boundaries_summary) +
  geom_sf(aes(fill = studied)) +
  coord_sf(xlim = c(-11, 30), ylim = c(34, 60)) +
  labs(title = "COVID Mortality, Europe", 
       fill = "Deaths In Period",
       caption = "Based off original map produced by Joel Smalley") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.line = element_blank(),
        panel.grid = element_blank(),
        plot.background = element_rect(fill = "#FAF7F2"),
        legend.background = element_blank())

```